### 2018-03-20 submission/manuscript.Rmd
discussion rewrite and reordering

In this study we show that the number of OTUs, error rate, and chimera prevalence are HiFi DNA polymerase dependent [Figure 1-2 & 4-5]. These differences are important because many diversity metrics rely on this measure as part of their calculations (e.g. richness). Additionally, the earlier detection of differences in total number of OTUs between HiFi DNA polymerases in the mock versus fecal samples might indicate that high biomass samples may underestimate the biases present within low biomass samples. Importantly, many of these differences are due to undetected chimeras that cannot be fully removed using standard bioinformatic approaches. This suggests that some of the diversity differences between studies can be attributed to differences in HiFi DNA polyermase used.

There were few differences in bacterial community composition based on HiFi DNA polymerase. Additionally, the data suggests that there are no differences in the overall bacterial community composition for sub-sampling depth or cycle number used. One possible reason for this outcome is that our study did not have enough power to detect differences due to low sample number in each group. Another reason is that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed HiFi DNA polymerase dependent differences in chimera prevalence.

@@

@@

Heavy standardization has been commonly suggested as a reasonable answer to finding the most reproducible approach. However, bias can be easily reproducible and can be found in every step of the 16S rRNA gene sequencing workflow. This study shows that the observed microbial community consistently varies based on HiFi DNA polymerase. Standardizing multiple workflows to one specific HiFi DNA polymerase could be detrimental since the PCR step not only misses entire species based on primer chosen () but also can artifically increase the number of OTUs observed. Arguably, standardizing workflow needs to be approached on a study by study basis and not every project needs to use the exact same approach. All aspects of the 16S rRNA gene sequencing workflow need to be customized for the specific microbial community that will be sampled. Although a diversity of approaches may make reproducibility more difficult it will help to avoid systematic biases from occuring due to widespread standardization of approaches.

@@

The number of OTUs are dependent on both HiFi DNA polymerase and cycle number chosen. Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys because intrinsic differences in HiFi DNA polymerases can change the number of OTUs observed and influence diversity based metrics that do not down-weight rare observations.

>>>>

In this study we show that the number of OTUs, error rate, and chimera prevalence are HiFi DNA polymerase dependent [Figure 1-2 & 4-5]. These differences are important because many diversity metrics rely on the number of OTUs or other measures dependent on error rate and chimera prevalence as part of their metric calculations (e.g. richness). Importantly, many of these differences are due to undetected chimeras that cannot be fully removed using standard bioinformatic approaches. This suggests that some of the diversity differences between studies can be attributed to differences in HiFi DNA polyermase used. Interestingly, the earlier detection of differences in total number of OTUs between HiFi DNA polymerases in the mock versus fecal samples might indicate that high biomass samples may underestimate the biases present within low biomass samples.

@@

There were few differences in bacterial community composition based on HiFi DNA polymerase. The data also suggests that there were no differences in the overall bacterial community composition for sub-sampling depth or cycle number used. One possible reason for this outcome was that our study did not have enough power to detect differences due to low sample number in each group. Another reason was that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed HiFi DNA polymerase dependent differences in chimera prevalence. Metrics that simply  look at presence/absence of OTUs (e.g. Jaccard [@Real1996]) may be less robust to chimera prevalence and by extension total number of OTU differences in HiFi DNA Polymerases. When choosing a distance metric careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account.

@@

Heavy standardization has been commonly suggested as a reasonable answer to finding the most reproducible approach. However, bias can be easily reproduced and can be found in every step of the 16S rRNA gene sequencing workflow. This study shows that specific diversity metrics used to measure the microbial community consistently vary based on HiFi DNA polymerase. Standardizing multiple workflows to one specific HiFi DNA polymerase could be detrimental since the PCR step not only misses entire species based on primer chosen [@Meisel2016] but also can artifically increase the number of OTUs observed. Arguably, the degree of workflow standardization across studies and research group needs to be approached on a study by study basis and not every project needs to use the exact same approach. All aspects of the 16S rRNA gene sequencing workflow need to be customized for the specific microbial community that will be sampled. Although a diversity of approaches may make reproducibility more difficult it will help to avoid systematic biases from occuring due to widespread standardization of approaches.

@@

The number of OTUs are dependent on both HiFi DNA polymerase and cycle number chosen. Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys because their intrinsic differences can change the number of OTUs observed and influence diversity based metrics that do not down-weight rare observations.

### 2018-03-20 submission/manuscript.tex
discussion rewrite and reordering

In this study we show that the number of OTUs, error rate, and chimera
prevalence are HiFi DNA polymerase dependent {[}Figure 1-2 \& 4-5{]}.
These differences are important because many diversity metrics rely on
this measure as part of their calculations (e.g.~richness).
Additionally, the earlier detection of differences in total number of
OTUs between HiFi DNA polymerases in the mock versus fecal samples might
indicate that high biomass samples may underestimate the biases present
within low biomass samples. Importantly, many of these differences are
due to undetected chimeras that cannot be fully removed using standard
bioinformatic approaches. This suggests that some of the diversity
differences between studies can be attributed to differences in HiFi DNA
polyermase used.

There were few differences in bacterial community composition based on
HiFi DNA polymerase. Additionally, the data suggests that there are no
differences in the overall bacterial community composition for
sub-sampling depth or cycle number used. One possible reason for this
outcome is that our study did not have enough power to detect
differences due to low sample number in each group. Another reason is
that many of the OTUs are likely not highly abundant, allowing the
Bray-Curtis index to be able to successfully down-weight chimeric OTUs
(Minchin, 1987). The choice of downstream diversity metric could be an
important consideration in helping to mitigate these observed HiFi DNA
polymerase dependent differences in chimera prevalence.

@@

Similar to using different perservation methods or different DNA
extraction kits, the type of HiFi DNA polymerase can add bias to the
observed bacterial community. The sequence error introduced by the HiFi
DNA polymerase is small and likely to be smaller than the biological
variation within a specific study, which would be consistent with
previous findings for preservation and DNA extraction methods (Salter et
al., 2014; Song et al., 2016; Luo et al., 2016). However, the chimarea
prevalence for some HiFi DNA polymerases (e.g.~Platinum) are relatively
large and might be greater than the oberseved biological variation
within a specific study. Within the larger context of the different 16S
rRNA gene sequencing steps, the choice of HiFi DNA polymerase can be as
important a consideration as either preservation or DNA extraction
method used.

Heavy standardization has been commonly suggested as a reasonable answer
to finding the most reproducible approach. However, bias can be easily
reproducible and can be found in every step of the 16S rRNA gene
sequencing workflow. This study shows that the observed microbial
community consistently varies based on HiFi DNA polymerase.
Standardizing multiple workflows to one specific HiFi DNA polymerase
could be detrimental since the PCR step not only misses entire species
based on primer chosen () but also can artifically increase the number
of OTUs observed. Arguably, standardizing workflow needs to be
approached on a study by study basis and not every project needs to use
the exact same approach. All aspects of the 16S rRNA gene sequencing
workflow need to be customized for the specific microbial community that
will be sampled. Although a diversity of approaches may make
reproducibility more difficult it will help to avoid systematic biases
from occuring due to widespread standardization of approaches.

@@

>>>>

In this study we show that the number of OTUs, error rate, and chimera
prevalence are HiFi DNA polymerase dependent {[}Figure 1-2 \& 4-5{]}.
These differences are important because many diversity metrics rely on
the number of OTUs or other measures dependent on error rate and chimera
prevalence as part of their metric calculations (e.g.~richness).
Importantly, many of these differences are due to undetected chimeras
that cannot be fully removed using standard bioinformatic approaches.
This suggests that some of the diversity differences between studies can
be attributed to differences in HiFi DNA polyermase used. Interestingly,
the earlier detection of differences in total number of OTUs between
HiFi DNA polymerases in the mock versus fecal samples might indicate
that high biomass samples may underestimate the biases present within
low biomass samples.

@@

There were few differences in bacterial community composition based on
HiFi DNA polymerase. The data also suggests that there were no
differences in the overall bacterial community composition for
sub-sampling depth or cycle number used. One possible reason for this
outcome was that our study did not have enough power to detect
differences due to low sample number in each group. Another reason was
that many of the OTUs are likely not highly abundant, allowing the
Bray-Curtis index to be able to successfully down-weight chimeric OTUs
(Minchin, 1987). The choice of downstream diversity metric could be an
important consideration in helping to mitigate these observed HiFi DNA
polymerase dependent differences in chimera prevalence. Metrics that
simply look at presence/absence of OTUs (e.g.~Jaccard (Real \& Vargas,
1996)) may be less robust to chimera prevalence and by extension total
number of OTU differences in HiFi DNA Polymerases. When choosing a
distance metric careful consideration of the biases introduced from the
PCR step of the 16S rRNA gene sequencing workflow need to be taken into
account.

Similar to using different preservation methods or different DNA
extraction kits, the type of HiFi DNA polymerase can add bias to the
observed bacterial community. The sequence error introduced by the HiFi
DNA polymerase is small and likely to be smaller than the biological
variation within a specific study, which would be consistent with
previous findings for preservation and DNA extraction methods (Salter et
al., 2014; Song et al., 2016; Luo et al., 2016). However, the chimarea
prevalence for some HiFi DNA polymerases (e.g.~Platinum) are relatively
large and might be greater than the oberseved biological variation
within a specific study. Within the larger context of the different 16S
rRNA gene sequencing steps, the choice of HiFi DNA polymerase can be as
important a consideration as either preservation or DNA extraction
method used.

Heavy standardization has been commonly suggested as a reasonable answer
to finding the most reproducible approach. However, bias can be easily
reproduced and can be found in every step of the 16S rRNA gene
sequencing workflow. This study shows that specific diversity metrics
used to measure the microbial community consistently vary based on HiFi
DNA polymerase. Standardizing multiple workflows to one specific HiFi
DNA polymerase could be detrimental since the PCR step not only misses
entire species based on primer chosen (Meisel et al., 2016) but also can
artifically increase the number of OTUs observed. Arguably, the degree
of workflow standardization across studies and research group needs to
be approached on a study by study basis and not every project needs to
use the exact same approach. All aspects of the 16S rRNA gene sequencing
workflow need to be customized for the specific microbial community that
will be sampled. Although a diversity of approaches may make
reproducibility more difficult it will help to avoid systematic biases
from occuring due to widespread standardization of approaches.

@@

\hypertarget{ref-Real1996}{}
Real R., Vargas JM. 1996. The probabilistic basis of jaccards index of
similarity. \emph{Systematic Biology} 45:380--385. DOI:
\href{https://doi.org/10.1093/sysbio/45.3.380}{10.1093/sysbio/45.3.380}.
